---
title: "RNA-seq Analysis in R"
subtitle: "Annotation and Visualisation of RNA-seq results - Solutions"
author: "Stephane Ballereau, Dominique-Laurent Couturier, Mark Dunning, Abbi Edwards, Ashley Sawle"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_document:
    toc: yes
    toc_float: yes
minutes: 300
layout: page
---

```{r setup, include=FALSE}
library(EnsDb.Mmusculus.v79)
library(DESeq2)
library(tidyverse)
```

```{r prepareData, echo=FALSE, message=FALSE, warning=FALSE}
# First load data and annotations
load("../Robjects/DE.RData")
load("../Robjects/Ensembl_annotations.RData")
```

## Exercise 1 - Retrieve the full annotation

> So far we have retrieved the annotation for just 1000 genes, but we need 
> annotations for the entire results table.
>
> A reminder of the code we have used so far:

```{r eval=FALSE}
# First load data if you haven't already
load("Robjects/DE.RData")

# lets set it up
ourCols <- c("GENEID", "SYMBOL", "ENTREZID")
ourKeys <- rownames(resLvV)[1:1000]

# run the query
annot <- AnnotationDbi::select(EnsDb.Mmusculus.v79, 
                keys=ourKeys, 
                columns=ourCols, 
                keytype="GENEID")
```

> (a) Run the same query using all of the genes in our results table (`resLvV`),
> and this time include the biotype of the genes too. Hint: You can find the 
> name of the column for this by running `columns(EnsDb.Mmusculus.v79)`
>
> (b) How many Ensembl genes have multipe Entrez IDs associated with them?
>
> (c) Are all of the Ensembl gene IDs annotated? If not, why do you think this 
> is?

```{r solutionChallenge1code, message=FALSE, warnings=FALSE}

# (a)

# use all of the genes
ourKeys <- rownames(resLvV)

# Include the biotype
columns(EnsDb.Mmusculus.v79)

## column 6 is "GENEBIOTYPE"
ourCols <- c("GENEID", "SYMBOL", "ENTREZID", "GENEBIOTYPE")

# run the query
annot <- AnnotationDbi::select(EnsDb.Mmusculus.v79, 
                keys=ourKeys, 
                columns=ourCols,
                keytype="GENEID")

# (b)
annot %>%  
    add_count(GENEID) %>%  
    dplyr::filter(n>1) %>% 
    distinct(GENEID) %>%
    count()

# (c)
dim(annot)
length(ourKeys)
```

> ### Challenge 2 {.challenge}

If you haven't already make sure you load in our data and annotation. Then shrink the values. You can copy and paste the code below.

```{r, echo=FALSE}
#Shrink our values
ddsShrink <- lfcShrink(ddsObj, coef="Status_lactate_vs_virgin")
shrinkLvV <- as.data.frame(ddsShrink) %>%
    rownames_to_column("GeneID") %>% 
    left_join(ensemblAnnot, "GeneID") %>% 
    rename(logFC=log2FoldChange, FDR=padj)
```

```{r, eval=FALSE}
# First load data and annotations if you haven't already
load("Robjects/DE.RData")
load("Robjects/Ensembl_annotations.RData")

#Shrink our values
ddsShrink <- lfcShrink(ddsObj, coef="Status_lactate_vs_virgin")
shrinkLvV <- as.data.frame(ddsShrink) %>%
    rownames_to_column("GeneID") %>% 
    left_join(ensemblAnnot, "GeneID") %>% 
    rename(logFC=log2FoldChange, FDR=padj)
```

> Use the log2 fold change (`logFC`) on the x-axis, and use `-log10(pvalue)` on the y-axis.
> (This `-log10` transformation is commonly used for p-values as it means that more
> significant genes have a higher scale) 
>
> (a) Create a new column of -log10(pvalue) values in shrinkLvV
>
> (b) Create a plot with points coloured by FDR < 0.05 similar to how we did in the MA plot


```{r volcanoPlot, fig.height=5, fig.width=7}
shrinkLvV %>% 
    mutate(`-log10(pvalue)` = -log10(pvalue)) %>% 
    ggplot(aes(x = logFC, y=`-log10(pvalue)`)) + 
      geom_point(aes(colour=FDR < 0.05), size=1)
```


