---
title: "RNA-seq Analysis in R"
subtitle: "Annotation and Visualisation of RNA-seq results - Solutions"
author: "Stephane Ballereau, Dominique-Laurent Couturier, Mark Dunning, Abbi Edwards, Ashley Sawle"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_document:
    toc: yes
    toc_float: yes
minutes: 300
layout: page
---

```{r setup, include=FALSE}
library(biomaRt)
library(GenomicFeatures)
library(DESeq2)
library(tidyverse)
```

```{r prepareData, message=FALSE, warning=FALSE}
# load data
load("../../Course_Materials/Robjects/DE.RData")
## set up connection to ensembl database
ensembl <- useEnsembl("genes")
# specify a data set to use
ensembl <- useDataset("mmusculus_gene_ensembl", mart=ensembl)
```

### Challenge 1

That was just 1000 genes. We need annotations for the entire results table.
Also, there may be some other interesting columns in BioMart that we wish to
retrieve.  

> (a) Search the attributes and add the following to our list of attributes:  
>       (i) The gene description   
>       (ii) The gene biotype    

```{r solutionChallenge1, message=FALSE, warnings=FALSE}
# check the available "attributes" - things you can retreive
listAttributes(ensembl) %>%
    filter(str_detect(name, "description"))

listAttributes(ensembl) %>%
    filter(str_detect(name, "biotype"))
```


> (b) Query BioMart using all of the genes in our results table (`resLvV`)

```{r solutionChallenge1bcode, message=FALSE, warnings=FALSE}
# set attributes
attributeNames <- c('ensembl_gene_id',
                    'entrezgene_id',
                    'external_gene_name',
                    'description',
                    'gene_biotype')

# Set the filter type and values
ourFilterType <- "ensembl_gene_id"

# set the values for the filter
filterValues <- rownames(resLvV)

# run the query
annot <- getBM(attributes=attributeNames,
               filters = ourFilterType,
               values = filterValues,
               mart = ensembl)
head(annot)
```
> (c) How many Ensembl genes have multipe Entrez IDs associated with them?  

```{r solutionChallenge1c, message=FALSE, warnings=FALSE}
# duplicate ids
annot %>%
  add_count(ensembl_gene_id) %>% 
  filter(n>1) %>% 
  distinct(ensembl_gene_id) %>% 
  nrow()
```

> (d) How many Ensembl genes in `resLvV` don't have any annotation?

```{r solutionChallenge1d, message=FALSE, warnings=FALSE}
length(filterValues) - length(unique(annot$ensembl_gene_id))
```

**Note**: The answers to (c) and (d) may change depending on the current release
version of BioMart. This html was generated with release 101.