---
title: "RNA-seq Analysis in R"
subtitle: "Annotation Differential Expression Results with biomaRt"
author: "Stephane Ballereau, Dominique-Laurent Couturier, Abbi Edwards, Ashley Sawle"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_document:
    toc: yes
minutes: 300
layout: page
bibliography: ref.bib
editor_options: 
  chunk_output_type: inline
---

```{r setup, message=FALSE}
library(biomaRt)
library(DESeq2)
library(tidyverse)
```

Before starting this section, we will make sure we have all the relevant objects
from the Differential Expression analysis.

```{r eval=FALSE}
load("Robjects/DE.RData")
```

```{r echo=FALSE}
load("../Course_Materials/Robjects/DE.RData")
```

# Overview

- Getting annotation

# Adding annotation to the DESeq2 results

We have a list of significantly differentially expressed genes, but the only
annotation we can see is the Ensembl Gene ID, which is not very informative. 

There are a number of ways to add annotation. One method is to do this using the
*org.Mm.eg.db* package. This package is one of several *organism-level* packages
which are re-built every 6 months. These packages are listed on the [annotation 
section](http://bioconductor.org/packages/release/BiocViews.html#___AnnotationData) 
of the Bioconductor, and are installed in the same way as regular Bioconductor 
packages. 

An alternative approach is to use `biomaRt`, an interface to the 
[BioMart](http://www.biomart.org/) resource. This is the method we will use 
today.

## Select BioMart database and dataset

The first step is to select the Biomart database we are going to access and 
which data set we are going to use.

There are multiple mirror sites that we could use for access. The default is to 
use the European servers, however if the server is busy or inaccessible for some
reason it is possible to access one of the three mirror sites. See the instructions
at [here](https://www.ensembl.org/info/data/biomart/biomart_r_package.html) for detailed
instruction on using different mirrors, but in brief simply add the `host` argument to
the `listMarts` and `useMart` functions below.   

e.g to use the US West mirror:  
  `ensembl=useMart("ENSEMBL_MART_ENSEMBL", host="uswest.ensembl.org")`  

# list the available datasets (species)

```{r connect}
# view the available databases
listMarts()
## set up connection to ensembl database
ensembl=useMart("ENSEMBL_MART_ENSEMBL")

# list the available datasets (species)
listDatasets(ensembl) %>% 
    filter(str_detect(description, "Mouse"))

# specify a data set to use
ensembl = useDataset("mmusculus_gene_ensembl", mart=ensembl)
```

## Query the database

Now we need to set up a query. For this we need to specify three things: 

(a) What type of information we are going to search the dataset on - called
**filters**. In our case this is Ensembl Gene IDs
(b) A vector of the **values** for our filter - the Ensembl Gene IDs from our DE 
results table
(c) What columns (**attributes**) of the dataset we want returned.

Returning data from Biomart can take time, so it's always a good idea to test 
your query on a small list of values first to make sure it is doing what you
want. We'll just use the first 1000 genes for now.

```{r queryBioMart, message=F}

# check the available "filters" - things you can filter for
listFilters(ensembl) %>% 
    filter(str_detect(name, "ensembl"))
# Set the filter type and values
ourFilterType <- "ensembl_gene_id"
filterValues <- rownames(resLvV)[1:1000]

# check the available "attributes" - things you can retreive
listAttributes(ensembl) %>% 
    head(20)
# Set the list of attributes
attributeNames <- c('ensembl_gene_id', 'entrezgene_id', 'external_gene_name')

# run the query
annot <- getBM(attributes=attributeNames, 
               filters = ourFilterType, 
               values = filterValues, 
               mart = ensembl)
```


### One-to-many relationships

Let's inspect the annotation.

```{r inspectAnnot}
head(annot)
dim(annot) # why are there more than 1000 rows?
length(unique(annot$ensembl_gene_id))

# find all rows containing duplicated ensembl ids
annot %>%  
    add_count(ensembl_gene_id) %>%  
    filter(n>1)
```

There are a couple of genes that have multiple entries in the retrieved 
annotation. This is becaues there are multiple Entrez IDs for a single Ensembl 
gene. These one-to-many relationships come up frequently in genomic databases, 
it is important to be aware of them and check when necessary. 

We will need to do a little work before adding the annotation to out results 
table. We could decide to discard one or both of the Entrez ID mappings, or we 
could concatenate the Entrez IDs so that we don't lose information. 

## Retrieve full annotation

> ### Challenge 1 {.challenge}
> That was just 1000 genes. We need annotations for the entire results table.
> Also, there may be some other interesting columns in BioMart that we wish to
> retrieve.  
>
> (a) Search the attributes and add the following to our list of attributes:  
>       (i) The gene description   
>       (ii) The gene biotype  
> (b) Query BioMart using all of the genes in our results table (`resLvV`)  
>
> (c) How many Ensembl genes have multipe Entrez IDs associated with them?  
> (d) How many Ensembl genes in `resLvV` don't have any annotation? Why is this?

